## Convert some numerical columns to categorical variables

Categorical variables are coded on NDA as numbers even if they have been collected as categorical variables in ABCD's database. The following steps will convert the numerical columns back into factor variables. This will matter in statistical models as factors have to be treated separately from continuous variables.

Read in the merged data (see [merge_data](notebooks/general/merge_data.md)) and a table that contains the factors for each categorical variable. The second table is stored in a csv file that is part of this repository and has been generated from the original REDCap data dictionaries.

```{r}
nda18 = readRDS("nda18_music.Rds")
categories = read.csv('choices_coding_nda18.csv')
```

This loop will go throught the column names stored in the categories table and convert those columns in the nda18 data frame to factor variables. However, we won't replace the numeric values with strings. AnalyzeActivities_all_2.0 expects activitiy values to be numeric, and that will be handled there

```{r}
# test this with su_tlfb_alc_sip (should not have a factor level for NA)
for (kitty in categories$name) {
    if (!(kitty %in% names(nda18))) next
    nda18[[kitty]] = factor(nda18[[kitty]])
    nda18[[kitty]][nda18[[kitty]] == ""] = NA
}
nda18 = droplevels(nda18)
```


If we ignore knows categorical variables we can try to convert all other variables to numeric variables if their levels are numerical:

```{r}
ncols = ncol(nda18)
colnames = names(nda18)
data_clean = nda18
typevec = NA
nlevvec = rep(NA,length(typevec))
is.wholenumber <- function(x, tol = .Machine$double.eps^0.5)  abs(x - round(x)) < tol | is.na(x)
for (coli in 3:ncols) {
  levvec = levels(as.factor(as.character(nda18[,coli])))
  nlev = length(levvec)
  levvec.numeric = suppressWarnings(as.numeric(levvec))
  nnum = sum((!is.na(levvec.numeric))|(levvec=="")|(levvec=="NA")) 
  nempty = sum(levvec==""|(levvec=="NA"))
  nlevvec[coli] = nlev
  if (names(nda18)[coli] %in% categories$name) {
      typevec[coli] = 'Categorical'
  } else if (nnum==nlev) { # All numeric
    data_clean[,coli] = as.numeric(as.character(nda18[,coli]))
    nint = sum(is.wholenumber(levvec.numeric))
    if (nint==nlev) {
      typevec[coli] = 'Integer'
    } else {
      typevec[coli] = 'Real'
    }
  } else if ((nnum-nempty)==0) { # No numeric, other than empty string
    if (nlev==2) {
      typevec[coli] = 'Binary'
    } else {
      typevec[coli] = 'Categorical'
    }
  } else {
    typevec[coli] = 'Ambiguous' # Inspect more closely
  }
  cat(sprintf('%5d: type=%s nlev=%d (%s)\n',coli,typevec[coli],nlevvec[coli],colnames[coli]))
}
nda18 = data_clean
# Ambiguius columns
colnames[typevec=='Ambiguous']

# Empty columns
#colnames[(typevec=='Integer')&(nlevvec==0)]

# All-zero columns
#colnames[(typevec=='Integer')&(nlevvec==1)]
#which((typevec=='Integer')&(nlevvec==1))
```

We can save this new version of the ABCD data combined spreadsheet now:
```{r}
saveRDS(nda18, "nda18_music.Rds")
```

Looking at the factor levels you will find that some of them are more difficult to read than others. Some contain HTML instructions for two language versions (English and Spanish) for each level. For dropdown entries the language encoding cannot use HTML, instead the following pattern is used: '##en##English##/en## ##es##Spanish##/es##'.
